Eurovision Song Contest: semifinal 1, 2009

### 2020-08-22 inst/eurovision.Rmd
bugfix and named object for eurovision

Each row corresponds to a competitor and each column to a judge.  Note
that the matrix is not square: the last two columns correspond to
Germany and the UK, who were judges but not competitors.  I have
removed the first column in the Wikipedia table (which gave the total
of the points), replaced self-voting entries with `NA`, and replaced
blanks with `0`.  Thus the first row corresponds to votes cast for
`ME`, Montenegro.  We see that `BE` (Belgium) awarded them three
points, `am` (Armenia) five points, and so on.

@@

In the above, matrix `preference` records voters' first, second, third,
etc choice.  A zero entry means no points (nul punkte!) and `NA` means
that voter was forbidden from voting for that player (countries cannot
vote for themselves).

The competitors were the first 18 countries (the last two countries,
Germany and the UK, voted but did not compete).  We need the voters'
choices to be the *rows* for consistency with the rest of the package
(and, for that matter, the aylmer package and indeed the emulator
package), so we take the transpose:

```{r,taketranspose}
preference <- t(preference)
preference
```

Now, take the first row of `preference`.  This represents the votes
cast *by* (sic) Montenegro (`"ME"`).  Their favourite was [last
column] Bosnia & Herzegovina (`"BA"`), who they gave rank 1 to.  Their
second favourite was Macedonia, their third was Turkey, and so on.
They were not allowed to vote for themselves, which is why the first
column is `NA`.  So the order was: `BA` (1), `MK` (2) , `TR` (3), `IS`
(4), `RO` (5), `IL` (6), `AM` (7), `FI` (8), `BY` (9), `MT` (10).  The
other countries (`CZ`, `BE`, `SW`, `CH`, `BG`, `PT`) did not receive
any points from Montenegro.  The rows show the votes cast by each
country; note the final two rows corresponding to votes cast by
Germany and the UK, who did not compete and therefore cast votes for
all competitors.

@@

```{r,morestuff}
eurovision2009 <- hyper2(d=18)

for(i in seq_len(nrow(preference))){   # cycle through the rows; each row is a voter
    d <- preference[i,,drop=TRUE]
    d[is.na(d)] <- -1  # kludge: make the voting country ineligible to vote.
    while(any(d>0)){
        eligible <- which(d>=0)   # This is why we set NA values to -1
        eurovision2009[which(d==1)] %<>% "+"(1)
                                            # The first choice among
                                             # eligible players has +1
                                              # power on the numerator

        eurovision2009[eligible] %<>% dec()   # denominator of all eligible players

        d[d==1] <- -1  # once you've won, you are ineligible to be chosen again
                       # NB, pipe idiom is ugly:  d %<>% `[<-`(. == 1, -1)

        d[d>0] %<>% dec()  # everyone moves down the list, so who
                            # *was* second choice becomes first
                            # choice, who *was* third choice becomes
                            # second, and so on.

    } # while() loop closes
} # i loop closes

## syntatic sugar:
pnames(eurovision2009) <- competitors
```

@@

```{r,checknull}
m <- maxp(eurovision2009)  # MLE
print(loglik(indep(m),eurovision2009))
print(loglik(indep(equalp(eurovision2009)),eurovision2009))
```

@@

```{r,orderplot}
points_scored <- rowSums(wiki_matrix,na.rm=TRUE)  # points
ox <- rank(-points_scored)
oy <- rank(-m)   # m = MLE
par(pty='s') # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,18),ylim=c(0,18),pch=16,
xlab="official order",ylab="my order",main="Eurovision 2009")
par(xpd=TRUE) # allow country names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[ox[i]],pos=4,col='gray') }
par(xpd=FALSE) # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

In the graph above, we see that both systems agree on the first four places.
We can also plot the numerical values:

```{r}
plot(log(rowSums(wiki_matrix,na.rm=TRUE)),log(m),
     pch=16,xlab="points scored",ylab="MLE strength")
```

>>>>

Each row corresponds to a contestant and each column to a judge.  Note
that the matrix is not square: the last two columns correspond to
Germany and the UK, who were judges but not competitors.  I have
removed the first column from the Wikipedia table (which gave the
total of the points), replaced self-voting entries with `NA`, and
replaced blanks with `0`.  Thus the first row corresponds to votes
cast for `ME`, Montenegro.  We see that `BY` (Belarus) awarded them
three points, `AM` (Armenia) five points, and so on.  The first column
corresponds to the points awarded by Monetenegro.  We see that they
placed gave Belarus (`BY`) two points, Armenia (`AM`) four points, and
so on.  Their favourite was Bosnia and Herzegovina (`BA`) to whom they
gave twelve points.  The points system used was:

@@

In the above, matrix `preference` records voters' first, second,
third, etc choice.  A zero entry means no points (nul punkte!) and
`NA` means that voter was forbidden from voting for that player
(countries cannot vote for themselves).  The competitors were the
first 18 countries (the last two countries,
Germany and the UK, voted but did not compete).

```{r}
preference
```
Now, take the first column of `preference`.  This represents the order
of preferences of Montenegro (`ME`).  Their favourite was [last row]
Bosnia & Herzegovina (`BA`), who they gave rank 1 to (that is, 12
points).  Their second favourite was Macedonia (`MK`), their third was
Turkey (`TR`), and so on.  They were not allowed to vote for themselves,
which is why the first row is `NA`.  So the order was: `BA` (first),
`MK` (second) , `TR` (third), `IS` (fourth), `RO` (fifth), `IL`
(sixth), `AM` (seventh), `FI` (eigthth), `BY` (ninth), `MT` (tenth).
The other countries (`CZ`, `BE`, `SW`, `CH`, `BG`, `PT`) came joint
last and did not receive any points from Montenegro.  Note the final
two rows corresponding to votes cast by Germany and the UK, who did
not compete and therefore cast votes for all competitors.

@@

```{r,morestuff}
eurovision2009 <- hyper2()

for(i in seq_len(ncol(preference))){   # cycle through the rows; each row is a voter
    d <- preference[,i,drop=TRUE]
	eurovision2009 <- 	eurovision2009  + ordervec2supp(d[!is.na(d)])
} # i loop closes
```

@@

```{r,checknull,cache=TRUE}
m <- maxp(eurovision2009)
m
equalp.test(eurovision2009)
```

@@

```{r,orderplot}
points_scored <- rowSums(wiki_matrix,na.rm=TRUE)  # points
ox <- rank(-points_scored)
oy <- rank(-m)   # m = MLE
oyp <- ordertrans(oy,names(ox))

par(pty='s') # square plot
plot(ox,oyp,asp=1,pty='s',xlim=c(0,18),ylim=c(0,18),pch=16,
xlab="official order",ylab="my order",main="Eurovision 2009")

par(xpd=TRUE) # allow country names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oyp[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE) # stop diagonal line from protruding beyond plotting region
abline(0,1)
```
